A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-14 tidyindex.qmd
update the gggi plot

To understand the uncertainty of this dimension reduction step while avoiding the missingness issue on the variable level, we can run a local tour to slightly vary the weight of each dimension to see how index value and country ranking changes. We select countries in the South Asia and Sub-Saharan Africa region and gradually increase the weight of one variable and reduce back to equal weight all four dimensions, one at a time, to produce an animation of how GGGI changes in each country. Five frames (equal weights and one for each dimension with a relatively higher weight) selected from the tour animation are shown in @fig-idx-tour and you can find the link to the full animation in figure caption.

Many insights can be derived from the animation and these selected frames. These can be helpful in understanding the sensitivity of the index value and country ranking to each variable and identifying the strengths and weaknesses of each country relative to others. When economy and education are given a greater weight, the variation in index values and ranking highlights countries that performs relatively better (moving right) or worse (moving left) in these two single dimensions. For example, the index moves to the left for Ethiopia and Senegal when economy is given a higher weight. This reveals the economy as a weakness for these two countries compared to their similarly-ranked peers. When health receives a higher weight, the impact on the index value is minimal. Notably, increasing the weight in politics leads to a pronounced decrease of index value for almost all the countries. Given the index value has a direct interpretation on how much of the gender gap has been closed, such a weight variation can cast a negative light on the progression to gender parity. Analysts need to further investigate the construction of the politics dimension and consider whether this is desirable for the index.

```{r eval = FALSE}
dt <- tidyindex::gggi %>% 
  dplyr::select(country: political_empowerment) %>%
  filter(country %in% c("Namibia", "South Africa", "Bangladesh", "Botswana",
                        "Ethiopia", "Kenya", "Senegal", "Afghanistan")) %>%
  arrange(index) 
colnames(dt) <- c(colnames(dt)[1:4], "economy", "education", "health", "politics")

w_proj2idx <- function(w){w/sum(w)}
w_idx2proj <- function(w){w/(sqrt(sum(w^2))) %>% matrix(nrow = length(w))}
find_next <- function(matrix){
  start <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1)
  tourr:::step_angle(tourr:::geodesic_info(start, matrix), 0.4)
}
b <- array(dim = c(4, 1, 9))
b[,,c(1,3,5,7,9)] <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1) %>% w_idx2proj()
b[,,2] <- matrix(c(0.7, 0.1, 0.1, 0.1), ncol = 1) %>% find_next()
b[,,4] <- matrix(c(0.1, 0.7, 0.1, 0.1), ncol= 1) %>% find_next()
b[,,6] <- matrix(c(0.1, 0.1, 0.5, 0.1), ncol= 1) %>% find_next()
b[,,8] <- matrix(c(0.1, 0.1, 0.1, 0.5), ncol = 1) %>% find_next()

render(
  dt[,5:8], planned2_tour(b), dev = "png",
  display_idx(cex = 4, label_cex = 3.5, panel_height_ratio = c(3,1), 
              axis_cex = 2.5, weight_cex = 3, label_col = "grey70", 
              label = dt$country, abb_vars = FALSE),
  width = 900, height = 900, apf = 1/20, frames = 100,
  here::here("figures/idx-tour/idx-tour-%03d.png")
)

gifski::gifski(
  list.files(here::here("figures/idx-tour"), full.names = TRUE),
  gif_file = here::here("figures/idx-tour.gif"),
  delay = 0.15, width = 900, height = 900)

@@

```{r fig-idx-tour, fig.align='center', fig.height=8} 
#| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index. The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full animation is available at https://vimeo.com/847874016?share=copy."
frames <- c("001", "013", "037", "061", "085")
ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl, ncol = 2)  
```

>>>>

To understand the uncertainty of this dimension reduction step from combining four dimensions (Economy, Education, Health, and Politics), we run a local tour [TODO: reference to local tour] that varies the weight of each dimension slightly to observe how index value and country ranking changes. The local tour produces an animation that gradually increases the weight of one variable and reduces it back to equal weight, one at a time. A set of countries are selected for this exercise, which includes 1) the top four with GGGI > 0.85, 2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the bottom five countries with GGGI < 0.6. Five frames (equal weights and one for each dimension with a relatively higher weight) are selected from the local tour animation to show in @fig-idx-tour and the full animation is available at  https://vimeo.com/847874016?share=copy.

@fig-idx-tour helps to understand the how different dimension reduction alternatives affect the whole index distribution and individual countries.

When economy and education are given a greater weight, the variation in index values and ranking highlights countries that performs relatively better (moving right) or worse (moving left) in these two single dimensions. For example, the index moves to the left for Ethiopia and Senegal when economy is given a higher weight. This reveals the economy as a weakness for these two countries compared to their similarly-ranked peers.

When health receives a higher weight, the impact on the index value is minimal.

Notably, increasing the weight in politics leads to a pronounced decrease of index value for almost all the countries. Given the index value has a direct interpretation on how much of the gender gap has been closed, such a weight variation can cast a negative light on the progression to gender parity. Analysts need to further investigate the construction of the politics dimension and consider whether this is desirable for the index.

```{r eval = FALSE}
dt <- tidyindex::gggi %>% 
  dplyr::select(country: political_empowerment) %>%
  filter(between(index, 0.72, 0.729)  | index < 0.6 | index >= 0.85) %>%
  arrange(index) 
colnames(dt) <- c(colnames(dt)[1:4], "economy", "education", "health", "politics")

w_proj2idx <- function(w){w/sum(w)}
w_idx2proj <- function(w){w/(sqrt(sum(w^2))) %>% matrix(nrow = length(w))}
find_next <- function(matrix, angle = 0.4){
  start <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1)
  tourr:::step_angle(tourr:::geodesic_info(start, matrix), angle = angle)
}
b <- array(dim = c(4, 1, 3))
b[,,1] <- matrix(c(0.3, 0.3, 0.3, 0.1), ncol = 1) %>% find_next()
b[,,2] <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1) %>% w_idx2proj()
b[,,3] <- matrix(c(0.1, 0.1, 0.1, 0.7), ncol= 1) %>% find_next(angle = 0.6)

render(
  dt[,5:8], planned2_tour(b), dev = "png",
  display_idx(half_range = 2,cex = 4, label_cex = 3.5, col = "black",
              panel_height_ratio = c(4,1),
              axis_bar_lwd = c(rep(3, 3), 7),
              axis_bar_col = c(rep("#000000", 3), "red"),
              axis_bar_label_cex = c(rep(2.5, 3), 3),
              axis_bar_label_col = c(rep("#000000", 3), "red"), 
              axis_label_cex = 3, frame_x_pos = 0.16, 
              axis_var_cex = 2.5,
              label = dt$country, abb_vars = FALSE),
  width = 900, height = 1400, apf = 1/20, frames = 120,
  here::here("figures/idx-tour/idx-tour-%03d.png")
)

gifski::gifski(
  list.files(here::here("figures/idx-tour"), full.names = TRUE),
  gif_file = here::here("figures/idx-tour.gif"),
  delay = 0.15, width = 900, height = 1400)

@@

* Bangladesh uses politics to compensate economy
* increase economy -> 
* increase education -> 
* increase health -> shrink the range of index (bring top and bottom country closer)
* increase politics -> all the country has index value smaller

```{r fig-idx-tour, fig.align='center', fig.height=5} 
#| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index. The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full animation is available at https://vimeo.com/847874016?share=copy."  
#frames <- c("001", "013", "037", "061", "085", "109")
frames <- c("001", "006", "012", "018", "024", "029")
ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl, ncol = 3)  
```

<!-- These can be helpful in understanding the sensitivity of the index value and country ranking to each variable and identifying the strengths and weaknesses of each country relative to others.  -->

### 2023-08-14 tidyindex.tex
update the gggi plot

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

To understand the uncertainty of this dimension reduction step while
avoiding the missingness issue on the variable level, we can run a local
tour to slightly vary the weight of each dimension to see how index
value and country ranking changes. We select countries in the South Asia
and Sub-Saharan Africa region and gradually increase the weight of one
variable and reduce back to equal weight all four dimensions, one at a
time, to produce an animation of how GGGI changes in each country. Five
frames (equal weights and one for each dimension with a relatively
higher weight) selected from the tour animation are shown in
Figure~\ref{fig-idx-tour} and you can find the link to the full
animation in figure caption.

Many insights can be derived from the animation and these selected
frames. These can be helpful in understanding the sensitivity of the
index value and country ranking to each variable and identifying the
strengths and weaknesses of each country relative to others. When
economy and education are given a greater weight, the variation in index
values and ranking highlights countries that performs relatively better
(moving right) or worse (moving left) in these two single dimensions.
For example, the index moves to the left for Ethiopia and Senegal when
economy is given a higher weight. This reveals the economy as a weakness
for these two countries compared to their similarly-ranked peers. When
health receives a higher weight, the impact on the index value is
minimal. Notably, increasing the weight in politics leads to a
pronounced decrease of index value for almost all the countries. Given
the index value has a direct interpretation on how much of the gender
gap has been closed, such a weight variation can cast a negative light
on the progression to gender parity. Analysts need to further
investigate the construction of the politics dimension and consider
whether this is desirable for the index.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable, interior hidden]}{\end{tcolorbox}}\fi

@@

To understand the uncertainty of this dimension reduction step from
combining four dimensions (Economy, Education, Health, and Politics), we
run a local tour {[}TODO: reference to local tour{]} that varies the
weight of each dimension slightly to observe how index value and country
ranking changes. The local tour produces an animation that gradually
increases the weight of one variable and reduces it back to equal
weight, one at a time. A set of countries are selected for this
exercise, which includes 1) the top four with GGGI \textgreater{} 0.85,
2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama,
Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the
bottom five countries with GGGI \textless{} 0.6. Five frames (equal
weights and one for each dimension with a relatively higher weight) are
selected from the local tour animation to show in
Figure~\ref{fig-idx-tour} and the full animation is available at
https://vimeo.com/847874016?share=copy.

Figure~\ref{fig-idx-tour} helps to understand the how different
dimension reduction alternatives affect the whole index distribution and
individual countries.

When economy and education are given a greater weight, the variation in
index values and ranking highlights countries that performs relatively
better (moving right) or worse (moving left) in these two single
dimensions. For example, the index moves to the left for Ethiopia and
Senegal when economy is given a higher weight. This reveals the economy
as a weakness for these two countries compared to their similarly-ranked
peers.

When health receives a higher weight, the impact on the index value is
minimal.

Notably, increasing the weight in politics leads to a pronounced
decrease of index value for almost all the countries. Given the index
value has a direct interpretation on how much of the gender gap has been
closed, such a weight variation can cast a negative light on the
progression to gender parity. Analysts need to further investigate the
construction of the politics dimension and consider whether this is
desirable for the index.

\begin{itemize}
\tightlist
\item
  Bangladesh uses politics to compensate economy
\item
  increase economy -\textgreater{}
\item
  increase education -\textgreater{}
\item
  increase health -\textgreater{} shrink the range of index (bring top
  and bottom country closer)
\item
  increase politics -\textgreater{} all the country has index value
  smaller
\end{itemize}
